---
title: "EDA"
author: "Fatimah Niyas"
date: "2025-04-15"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploratory Data Analysis

### Loading necessary libraries
```{r}
library(tidyverse)
library(ggplot2)
library(maps)
library(dplyr)
```

### Loading in meteorite landings dataset

```{r}
setwd("~/DATA 205")
meteorite <- read.csv('Meteorite_Landings.csv')
```

### Data cleaning before analysis
```{r}
meteorite1 <- meteorite |>
  filter(!is.na(mass..g.)) |> # filter out any NAs
  filter(!is.na(year)) |>
  filter(!is.na(reclat)) |>
  filter(!is.na(reclong)) |>
  rename(mass = mass..g.)|> # renaming for easier access
  filter(year > 1850 & year < 2100) |> # filter the years to avoid big outliers
  filter(reclat >= -90 & reclat <= 90, reclong >= -180 & reclong <= 180) |> # make sure the coordinates are valid
  filter(mass > 0 & mass < 1e6) #filter out mass values
head(meteorite1)
```

### Visualization on meteorite count over years

This visualization can help answer the questions: 

Have meteorite landing findings increased over time?
What factors may have caused this?
Has improved technology led to more accurate detection and reporting?

```{r}
ggplot(meteorite1, aes(x = year)) + # visualizing meteorite count over years
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  labs(title = "Number of Meteorites Recorded(1850-2013)",
       x = "Year", y = "Count") +
  theme_minimal()
```
As we can see here, the data is skewed left, clustered in more recent years likely because of advancements in technology, increased global scientific interest, and improved tracking and reporting systems. In the past, many meteorite events may have gone unnoticed or undocumented, especially in remote or less-populated areas. As scientific tools have developed, more meteorite landings have been detected, recorded, and analyzed—leading to a sharp increase in entries during the 20th and 21st centuries.

### Visualization on meteorite mass over years

These visualizations can help answer the questions: 
Has improved technology led to more accurate detection and reporting?
Is there a relationship between meteorite mass and the year it was found?
Do technological improvements make smaller meteorites easier to detect now?

```{r}
mass_year <- meteorite1 |>
  filter(mass < 10000) # adjusting to visualize better

ggplot(mass_year, aes(x = year, y = mass)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Meteorite Mass Over Time",
       x = "Year", y = "Mass (g)")
```
Boxplot: 
```{r}
mass_year$decade <- floor(as.numeric(mass_year$year) / 10) * 10 # turning year into by decade

ggplot(mass_year, aes(x = factor(decade), y = mass)) +
  geom_boxplot(outlier.shape = NA, fill ="darkolivegreen") +
  scale_y_log10() +
  labs(title = "Distribution of Meteorite Mass by Decade",
       x = "Decade", y = "Mass (log scale)")
```

Here, we can see that in more recent years, the number of meteorites with smaller masses has increased also due to advancements in detection technology and improved reporting systems. Smaller meteorites that would have gone unnoticed in the past are now being recovered thanks to tools like metal detectors, satellite tracking, etc.

### Meteorite locations on a map

This visualization can help answer the questions: 
What locations have a higher amount of meteorite findings? 
Are meteorite discoveries clustered in specific countries or regions?
Are specific countries reporting more meteorite landings due to better detection technology?


```{r}
world_map <- map_data("world")

ggplot() +
  geom_map(data = world_map, map = world_map,
           aes(x = long, y = lat, map_id = region),
           fill = "lightgray", color = "white", size = 0.2) +
  geom_point(data = meteorite1,
             aes(x = reclong, y = reclat),
             color = "darkblue", alpha = 0.5, size = 1) +
  labs(title = "Meteorite Landings (After 1850s)",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

### Meteorite locations on a map colored by 'Fell' vs. 'Found'

```{r}
ggplot() +
  geom_map(data = world_map, map = world_map,
           aes(x = long, y = lat, map_id = region),
           fill = "lightgray", color = "white", size = 0.2) +
  geom_point(data = meteorite1,
             aes(x = reclong, y = reclat, color = fall),
             alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Fell" = "red", "Found" = "blue")) +
  labs(title = "Meteorite Landings (After 1850s)",
       x = "Longitude", y = "Latitude", color = "Fall Status") +
  theme_minimal()
```   


